Nodal Discontinuous Galerkin Method for the Two-Moment Model of Neutrino Transport

### 2017-06-01 Documents/M1/dgM1Curvilinear.tex
minor modifications to text

When coupling neutrinos to the background fluid, the gain/loss terms involve integrals of the right-hand sides in Eqs.~\eqref{eq:momentEquationEnergy} and \eqref{eq:momentEquationMomentum} over energy space.  
To simulate multi-group neutrino transport and compute the lepton and energy exchange with the fluid, the infinite energy domain $\bbR^{+}$ is replaced with a finite domain $D^{\epsilonNu}=\{\epsilonNu~|~\epsilonNu\in[0,\epsilonNu_{\mbox{\tiny max}}]\}$.  
The domain $D^{\epsilonNu}$ is then split up into $N_{\epsilonNu}$ energy bins $K_{i}^{\epsilonNu}=\{\epsilonNu~|~\epsilonNu\in[\epsilonNu_{i-1/2},\epsilonNu_{i+1/2}]\}$, where the width of the $i$th bin is $|K_{i}^{\epsilonNu}|=(\epsilonNu_{i+1/2}-\epsilonNu_{i-1/2})$.  
Within each bin, we employ a polynomial representation of the radiation fields and solve for nodal values collocated in Gaussian quadrature points.  
The number of quadrature points $N$ then increases with the degree $k$ of the polynomial representation ($N=k+1$).  
Integrals of radiation quantities over energy space are then evaluated as
\begin{equation}
  \int_{\bbR^{+}}g(\epsilonNu)\,d\epsilonNu
  \approx\int_{D^{\epsilonNu}}g(\epsilonNu)\,d\epsilonNu
  =\sum_{i=1}^{N^{\epsilonNu}}\int_{K_{i}^{\epsilonNu}}g(\epsilonNu)\,d\epsilonNu
  \approx\sum_{i=1}^{N^{\epsilonNu}}|K_{i}^{\epsilonNu}|\sum_{q=1}^{N}w_{q}\,g_{q}=\sum_{j=1}^{M^{\epsilonNu}} W_{j}\,g_{j}=\vect{W}\cdot\vect{g},
\end{equation}
where $w_{q}$ are Gaussian quadrature weights (normalized such that $\sum_{q=1}^{N}w_{q}=1$) and $M^{\epsilonNu}=N^{\epsilonNu}\times N$.  
In constructing the vectors $\vect{W},\vect{g}\in\bbR^{M^{\epsilonNu}}$, the elements are sorted with monotonically increasing energy; e.g., $W_{j}=|K_{i}^{\epsilonNu}|\,w_{q}$, where $j=(i-1)\times N+q$, ($i=1,\ldots,N^{\epsilonNu}$, $q=1,\ldots,N$).

@@

Electron capture on nucleons and nuclei results in changes to the electron fraction and internal energy of the fluid given by
\begin{align}
  d_{t}Y_{e}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\int_{\bbR^{+}}\tilde{\chi}\,\big(\cJ_{0}-\cJ\big)\,dV_{\epsilonNu}, \label{eq:electronFractionEmAb} \\
  d_{t}\epsilon
  &=-\f{1}{\rho}\int_{\bbR^{+}}\tilde{\chi}\,\big(\cJ_{0}-\cJ\big)\,\epsilonNu\,dV_{\epsilonNu}, \label{eq:internalEnergyEmAb}
\end{align}
where $m_{\mbox{\tiny B}}$ is the baryon mass and $\epsilon$ is the specific internal energy.  
In each implicit step in the SIRK2 scheme discussed in Section~\ref{sec:semiImplicitTime} we solve Eqs.~\eqref{eq:electronFractionEmAb}-\eqref{eq:internalEnergyEmAb} coupled to the collision part of the moment equations.  
\begin{align}
  Y_{e}^{n+1}-Y_{e}^{n}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(2)}\,\gamma_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n+1}\big), \label{eq:electronFractionEmAbDiscrete0} \\
  \epsilon^{n+1}-\epsilon^{n}
  &=-\f{1}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(3)}\,\gamma_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n+1}\big), \label{eq:internalEnergyEmAbDiscrete0}
\end{align}
where $W_{j}^{(2,3)}=w_{q}\,\epsilonNu_{i,q}^{2,3}\,\Delta\epsilonNu_{i}$ and $\gamma_{j}^{n}=\dt\,\tilde{\chi}_{j}^{n}$.  
Note that the absorption opacity is defined at the known time level $t^{n}$, which avoids computation of derivatives with respect to temperature and electron fraction.  
The radiation energy density is updated from
\begin{equation}
  \cJ_{j}^{n+1}=\f{\cJ_{j}^{n}+\gamma_{j}^{n}\,\cJ_{0,j}^{n+1}}{1+\gamma_{j}^{n}}, 
  \label{eq:momentEquationEnergyEmAbDiscrete}
\end{equation}
which, when inserted into Eqs.~\eqref{eq:electronFractionEmAbDiscrete0}-\eqref{eq:internalEnergyEmAbDiscrete0}, results in a nonlinear system for the unknowns $Y^{n+1}$ and $\epsilonNu^{n+1}$
\begin{align}
  Y_{e}^{n+1}-Y_{e}^{n}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(2)}\,\tilde{\gamma}_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n}\big), \label{eq:electronFractionEmAbDiscrete} \\
  \epsilon^{n+1}-\epsilon^{n}
  &=-\f{1}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(3)}\,\tilde{\gamma}_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n}\big), \label{eq:internalEnergyEmAbDiscrete}
\end{align}
where $\tilde{\gamma}_{j}^{n}=\gamma_{j}^{n}/(1+\gamma_{j}^{n})$.  
Since the equilibrium density $\cJ_{0,j}$ is a nonlinear function of $Y_{e}$ and $\epsilon$, Eqs.~\label{eq:electronFractionEmAbDiscrete}-\label{eq:internalEnergyEmAbDiscrete} are solved using Newton's method.  
Once $Y_{e}^{n+1}$ and $\epsilon^{n+1}$ are obtained, the radiation density is obtained from Eq.~\eqref{eq:momentEquationEnergyEmAbDiscrete}, while the flux is given by
\begin{equation}
  \bcH_{j}^{n+1}=\bcH_{j}^{n}/(1+\gamma_{j}^{n}).  
  \label{eq:momentEquationMomentumEmAbDiscrete}
\end{equation}

>>>>

When coupling neutrinos to the background fluid, the gain/loss terms involve integrals of the right-hand sides in Eqs.~\eqref{eq:momentEquationEnergy} and \eqref{eq:momentEquationMomentum} over energy space.  
To simulate multi-group neutrino transport and compute the lepton and energy exchange with the fluid, the infinite energy domain $\bbR^{+}$ is replaced with a finite domain $D^{\epsilonNu}=\{\epsilonNu~|~\epsilonNu\in[0,\epsilonNu_{\mbox{\tiny max}}]\}$.  
The domain $D^{\epsilonNu}$ is then split up into $N_{\epsilonNu}$ energy bins $K_{i}^{\epsilonNu}=\{\epsilonNu~|~\epsilonNu\in[\epsilonNu_{i-1/2},\epsilonNu_{i+1/2}]\}$, where the width of the $i$th bin is $|K_{i}^{\epsilonNu}|=(\epsilonNu_{i+1/2}-\epsilonNu_{i-1/2})$.  
Within each bin, we employ a polynomial representation of the radiation fields and solve for nodal values collocated in Gaussian quadrature points.  
The number of quadrature points $N$ then increases with the degree $k$ of the polynomial representation ($N=k+1$).  
Integrals of radiation quantities over energy space are then evaluated as
\begin{equation}
  \int_{\bbR^{+}}g(\epsilonNu)\,d\epsilonNu
  \approx\int_{D^{\epsilonNu}}g(\epsilonNu)\,d\epsilonNu
  =\sum_{i=1}^{N^{\epsilonNu}}\int_{K_{i}^{\epsilonNu}}g(\epsilonNu)\,d\epsilonNu
  \approx\sum_{i=1}^{N^{\epsilonNu}}|K_{i}^{\epsilonNu}|\sum_{q=1}^{N}w_{q}\,g_{q}=\sum_{j=1}^{M^{\epsilonNu}} W_{j}\,g_{j}=\vect{W}\cdot\vect{g},
\end{equation}
where $w_{q}$ are Gaussian quadrature weights (normalized such that $\sum_{q=1}^{N}w_{q}=1$) and $M^{\epsilonNu}=N^{\epsilonNu}\times N$.  
In constructing the vectors $\vect{W},\vect{g}\in\bbR^{M^{\epsilonNu}}$, their elements are sorted with energy monotonically increasing with $j$; e.g., $\epsilonNu_{j}=\epsilonNu_{i-1/2}+|K_{i}|\times(1/2+\eta_{q}^{\epsilonNu})$ and $W_{j}=|K_{i}^{\epsilonNu}|\,w_{q}$, where $\eta_{q}^{\epsilonNu}$ is the Gaussian quadrature point on the local reference element ($\eta^{\epsilonNu}\in[-1/2,1/2]$), and $j=(i-1)\times N+q$, ($i=1,\ldots,N^{\epsilonNu}$, $q=1,\ldots,N$).

@@

Electron capture on nucleons and nuclei results in changes to the electron fraction and internal energy of the fluid given by
\begin{align}
  d_{t}Y_{e}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\int_{\bbR^{+}}\tilde{\chi}\,\big(\cJ_{0}-\cJ\big)\,dV_{\epsilonNu}, \label{eq:electronFractionEmAb} \\
  d_{t}\epsilon
  &=-\f{1}{\rho}\int_{\bbR^{+}}\tilde{\chi}\,\big(\cJ_{0}-\cJ\big)\,\epsilonNu\,dV_{\epsilonNu}, \label{eq:internalEnergyEmAb}
\end{align}
where $m_{\mbox{\tiny B}}$ is the baryon mass and $\epsilon$ is the specific internal energy.  
In each implicit step in the SIRK2 scheme discussed in Section~\ref{sec:semiImplicitTime} we solve Eqs.~\eqref{eq:electronFractionEmAb}-\eqref{eq:internalEnergyEmAb} coupled to the collision part of the moment equations.  
\begin{align}
  Y_{e}^{n+1}-Y_{e}^{n}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(2)}\,\gamma_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n+1}\big), \label{eq:electronFractionEmAbDiscrete0} \\
  \epsilon^{n+1}-\epsilon^{n}
  &=-\f{1}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(3)}\,\gamma_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n+1}\big), \label{eq:internalEnergyEmAbDiscrete0}
\end{align}
where $W_{j}^{(2,3)}=w_{q}\,\epsilonNu_{i,q}^{2,3}\,\Delta\epsilonNu_{i}$ and $\gamma_{j}^{n}=\dt\,\tilde{\chi}_{j}^{n}$.  
Note that the absorption opacity is defined at the known time level $t^{n}$, which avoids computation of derivatives with respect to temperature and electron fraction.  
The radiation energy density is updated from
\begin{equation}
  \cJ_{j}^{n+1}=\f{\cJ_{j}^{n}+\gamma_{j}^{n}\,\cJ_{0,j}^{n+1}}{1+\gamma_{j}^{n}}, 
  \label{eq:momentEquationEnergyEmAbDiscrete}
\end{equation}
which, when inserted into Eqs.~\eqref{eq:electronFractionEmAbDiscrete0}-\eqref{eq:internalEnergyEmAbDiscrete0}, results in a nonlinear system for the unknowns $Y^{n+1}$ and $\epsilon^{n+1}$
\begin{align}
  Y_{e}^{n+1}-Y_{e}^{n}
  &=-\f{m_{\mbox{\tiny B}}}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(2)}\,\tilde{\gamma}_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n}\big), \label{eq:electronFractionEmAbDiscrete} \\
  \epsilon^{n+1}-\epsilon^{n}
  &=-\f{1}{\rho}\sum_{j=1}^{M^{\epsilonNu}}W_{j}^{(3)}\,\tilde{\gamma}_{j}^{n}\,\big(\cJ_{0,j}^{n+1}-\cJ_{j}^{n}\big), \label{eq:internalEnergyEmAbDiscrete}
\end{align}
where $\tilde{\gamma}_{j}^{n}=\gamma_{j}^{n}/(1+\gamma_{j}^{n})$.  
Since the equilibrium density $\cJ_{0,j}$ is a nonlinear function of $Y_{e}$ and $\epsilon$, Eqs.~\eqref{eq:electronFractionEmAbDiscrete}-\eqref{eq:internalEnergyEmAbDiscrete} are solved using Newton's method.  
Once $Y_{e}^{n+1}$ and $\epsilon^{n+1}$ are obtained, the radiation density is obtained from Eq.~\eqref{eq:momentEquationEnergyEmAbDiscrete}, while the flux is given by
\begin{equation}
  \bcH_{j}^{n+1}=\bcH_{j}^{n}/(1+\gamma_{j}^{n}).  
  \label{eq:momentEquationMomentumEmAbDiscrete}
\end{equation}
